---
title: "Global Coal Power & Emissions R Project"
author: "Logan Rosell"
format: pdf
editor: visual
---

```{r}

library(ggplot2)
library(tidyverse)
library(skimr)
library(car)
library(gtsummary)
library(gt)
library(lawstat)
library(sf)
library(rnaturalearth)
library(extrafont)
library(here)

#font_import() 
#loadfonts(device = "pdf")

theme_set(theme_classic())
theme_update(text = element_text(family = "Times New Roman"))

```

Data Source: https://ember-energy.org/data/yearly-electricity-data/

Question: Does coal power's CO2 emission intensity vary across global regions and time between 2000 and 2024?

# Cleaning Data

```{r}

global_power_data <- read_csv(here::here("data", "yearly_full_release_long_format.csv"))

```

```{r}
global_coal <- global_power_data %>%
  filter(Variable == 'Coal') %>%
  filter(Category != "Capacity") %>%
  dplyr::select(-c(`YoY absolute change`, `YoY % change`, Category, Subcategory, Variable)) %>%
  pivot_wider(names_from = Unit, values_from = Value) %>%
  mutate(mtCO2_per_TWh = mtCO2/TWh,
  mtCO2_per_TWh = ifelse(is.na(mtCO2_per_TWh), 0, mtCO2_per_TWh))
```

# Overview of Data

```{r}
skim(global_power_data)
summary(global_power_data)
```

```{r}
countries <- global_power_data %>% filter(`Area type`=="Country or economy") %>% dplyr::select(Area)
unique(countries)
```

# Region Analysis

```{r}
global_coal %>%
  filter(TWh != 0) %>%
  filter(`Area type` == "Country or economy") %>%
  ggplot(aes(x=Year, y=mtCO2_per_TWh)) +
  geom_point(size=0.5) +
  facet_wrap(~Continent) +
  theme(
    axis.text.x = element_text(
      angle = 45,    
      vjust = 1,
      hjust = 1
    )
  ) +
  labs(title="CO2 Intensity Variation Within Region Over Time")
ggsave(
  filename = here("outputs", "plots", "co2_intensity_variation_within_region_over_time.png"),
  width = 6, 
  height = 4, 
  units = "in", 
  dpi = 1500
)
```

```{r}
region_data <- global_coal %>%
  filter(`Area type` == 'Region') %>%
  filter(Area %in% c("Africa",
                      "Asia",
                      "Europe",
                      "Latin America and Caribbean",
                      "Middle East",
                      "North America",
                      "Oceania")) %>%
  dplyr::select(-c(`ISO 3 code`, Continent, `Ember region`, EU, OECD, G20, G7, ASEAN)) %>%
  mutate(percent_100 = `%` / 100)
```

```{r}
table_1 <- region_data %>%
  tbl_summary(by = Area,
              include = c(`%`, TWh, mtCO2, mtCO2_per_TWh),
              statistic = all_continuous() ~ "{mean} ({median})") %>%
  as_gt()
gtsave(data = table_1, filename = "summary_stats_table_by_region.png")

table_2 <- region_data %>%
  tbl_summary(by = Year,
              include = c(`%`, TWh, mtCO2, mtCO2_per_TWh),
              statistic = all_continuous() ~ "{mean} ({median})") %>%
  as_gt()
gtsave(
  data = table_2, 
  filename = here("outputs", "plots", "summary_stats_table_by_year.png")
)
```

```{r}
ggplot(region_data, aes(x=Year, y=mtCO2_per_TWh, color=Area)) +
  geom_line() +
  labs(title="CO2 Intensity Over Time by World Region",
       y= "mtCO2 per TWh")

ggsave(
  filename = here("outputs", "plots", "co2_intensity_over_time_by_world_region.png"),
  width = 6, 
  height=4,
  units = "in",
  dpi = 1500
)
```

## Summarizing Region Data

```{r}
skim(region_data)
summary(region_data)
```

```{r}
ggplot(region_data, aes(x=Area, y=mtCO2_per_TWh, color=Area)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
  labs(x="",
       title="Distribution of CO2 per TWh from Coal Power",
       color = "Region",
       y= "mtCO2 per TWh")
ggsave(
  filename = here("outputs", "plots", "distribution_of_co2_per_twh_from_coal_power.png"),
  width = 7,
  height=4,
  units = "in",
  dpi = 1500
)

```

### Summary Stats by Region

```{r}
region_data %>%
  group_by(Area) %>%
  summarize(
    N = n(),
    Mean = mean(mtCO2_per_TWh),
    Median = median(mtCO2_per_TWh),
    SD = sd(mtCO2_per_TWh),
    Min = min(mtCO2_per_TWh),
    Max = max(mtCO2_per_TWh)
  ) %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  gt()
```

```{r}
ggplot(region_data, aes(y=mtCO2_per_TWh)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1)) +
  labs(x="Region",
       title="Distribution of CO2 per TWh from Coal Power--Separate Scales",
       color = "Region",
       y= "mtCO2 per TWh") +
  facet_wrap(~ Area, nrow = 2, scales = "free_y")

ggsave(
  filename = here("outputs", "plots", "distribution_of_co2_per_twh_from_coal_power_separate_scales.png"),
  width = 7,
  height = 4,
  units = "in",
  dpi = 1500
)
```

```{r}
ggplot(region_data, aes(x=mtCO2_per_TWh, fill=Area)) +
  geom_histogram() +
  labs(title="Histogram of CO2 per TWh from Coal by Region",
       fill="Region",
       y="Count",
       x="mtCO2 per TWh")
ggsave(
  filename = here("outputs", "plots", "histogram_of_co2_per_twh_from_coal_by_region.png"),
  width = 7,
  height = 4,
  units = "in",
  dpi = 1500
)
```

# Testing the difference in CO2 Intensity between regions

Null Hypothesis B: Coal power CO2 emission intensity does NOT vary between global regions.

Alternative Hypothesis B: Coal power CO2 emission intensity DOES vary between global regions.

```{r}
shapiro.test(region_data$mtCO2_per_TWh)
```

A small p-value from our Shaprio-Wilks test tells use we should reject the null that our CO2_per_TWh data is normally distributed.

```{r}
levene.test(region_data$mtCO2_per_TWh, group=region_data$Area)
```

Small P-value for Brown-Forsythe means we reject the null of equal variance.

```{r}
kruskal.test(mtCO2_per_TWh ~ Area, data=region_data)
```

Because we don't have normally distributed data or equal variance I used the Kruskal-Wallis test (robust to both issues) and got a small p-value allowing us to reject the null that the distributions and medians are the same across global regions.

```{r}
pairwise.wilcox.test(region_data$mtCO2_per_TWh, g = region_data$Area)
```

All Pairs of regions are statically significantly different from each other.

# Mapping the data (2024)

```{r}
only_2024 <- global_coal %>%
  filter(Year == "2024") %>%
  filter(`Area type` == "Country or economy") %>%
  mutate(iso_3 = toupper(trimws(`ISO 3 code`))) %>%
  mutate(log_intensity = log((mtCO2_per_TWh * 100)))
```

```{r}
world_map <- ne_countries(scale = "medium", returnclass = "sf")

map_data_2024 <- world_map %>%
  left_join(only_2024, by = c("iso_a3" = "iso_3"))
```

```{r}
ggplot(data = map_data_2024) +
  geom_sf(aes(fill = log_intensity), 
          color = "gray80", 
          size = 0.1) +
  scale_fill_viridis_c(
    option = "viridis", 
    direction = -1,    
    name = "mtCO2 per TWh", 
    na.value = "lightgray",
    guide = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = 10,
      barheight = 0.5,
      direction = "horizontal")
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(vjust = 0.5)) +
  labs(
    title = "Global Coal Carbon Intensity by Country (2024)") +
  # No Antarctica
  coord_sf(ylim = c(-55, 85))

ggsave(
  filename = here("outputs", "plots", "global_coal_carbon_intensity_by_country_2024.png"),
  width = 6,
  height = 4,
  units = "in",
  dpi = 1500
)
```

# Modeling Intensity Over Time By Region

```{r}
intensity_time_region_model <- lm(formula = mtCO2_per_TWh ~ Year * Area, data=region_data)
summary(intensity_time_region_model)
shapiro.test(region_data$mtCO2_per_TWh)
  
  old_par <- par(no.readonly = TRUE)
  diagnostic_plot_title = "Global Diagnostic Plot"
  plot_file_name = paste0(diagnostic_plot_title, ".png")



  png(filename = plot_file_name,
    width = 1500,
    height = 1000,
    units = "px",
    res = 150)
  par(mfrow = c(2, 2), 
    family = "serif",
    oma = c(0, 0, 2, 0),
    mar = c(5, 5, 2, 1))
  plot(intensity_time_region_model, sub=" ")
  mtext(text = diagnostic_plot_title,
        outer = TRUE,
        cex = 1.5,
        font = 2,
        line = 0,
        adj = 0)
  dev.off()
  par(old_par)




```

```{r}
vif(intensity_time_region_model)
```

```{r}
ggplot(region_data, aes(x=Year, y=mtCO2_per_TWh, color=Area)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title="Modeling CO2 per TWh Over Time by Region",
       y= "mtCO2 per TWh")

ggsave(
  filename = here("outputs", "plots", "modeling_co2_per_twh_over_time_by_region.png"),
  width = 7,
  height = 4,
  units = "in",
  dpi = 1500
)
```

```{r}
region_model_maker_2 <- function(region) {
  df <- region_data %>%
  filter(Area == region)
  old_par <- par(no.readonly = TRUE)
  diagnostic_plot_title = paste(region, "Diagnostic Plot")
  plot_file_name = paste0(diagnostic_plot_title, ".png")

  model <- lm(formula = mtCO2_per_TWh ~ Year, data=df)
  print(summary(model))
  png(filename = plot_file_name,
    width = 1500,
    height = 1000,
    units = "px",
    res = 150)
  par(mfrow = c(2, 2), 
    family = "serif",
    oma = c(0, 0, 2, 0),
    mar = c(5, 5, 2, 1))
  plot(model, sub=" ")
  mtext(text = diagnostic_plot_title,
        outer = TRUE,
        cex = 1.5,
        font = 2,
        line = 0,
        adj = 0)
  dev.off()
  par(old_par)
  plot(ggplot(df, aes(x=Year, y=mtCO2_per_TWh)) + geom_point())
  plot(ggplot(df, aes(x=mtCO2_per_TWh))) + geom_histogram()

}
```

```{r}
regions = c("Asia",
            "Africa",
            "Europe",
            "Latin America and Caribbean",
            "Middle East",
            "Oceania",
            "North America")
for (i in regions){
   df <- region_data %>%
  filter(Area == i)
   print(i)
   print(shapiro.test(df$mtCO2_per_TWh))
}

```

```{r}
region_model_maker_2("Asia")
```

```{r}
region_model_maker_2("Africa")
```

```{r}
region_model_maker_2("Europe")
```

```{r}
region_model_maker_2("Latin America and Caribbean")
```

```{r}
region_model_maker_2("Middle East")
```

```{r}
region_model_maker_2("Oceania")
```

```{r}
region_model_maker_2("North America")
```

```{r}
good_models_regions <- c("Europe",
                         "Latin America and Caribbean",
                         "Asia",
                         "Oceania")
good_models_data <- region_data %>%
  filter(Area %in% good_models_regions)

ggplot(good_models_data, aes(x=Year, y=mtCO2_per_TWh)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(title="Models of Reasonable Fit",
       y = "mtCO2 per TWh") +
  facet_wrap(~Area, scales = "free_y") +
  theme(plot.margin = unit(c(5, 10, 5, 5), "mm"))

ggsave(
  filename = here("outputs", "plots", "models_of_reasonable_fit.png"),
  width = 7,
  height = 4,
  units = "in",
  dpi = 1500
)
```
